---
title: 'ML with cv.glmnet'
author: "Holly Giles, adapted from a script by Junyan Lu"
date: "`r doc_date()`"
output:
  BiocStyle::html_document:
      toc: yes
      toc_depth: 3
      toc_float: yes
      code_folding: "hide" 
---


This vignette is a short template script that can be adapted for running basic lasso-regularised regression. 

## Set up
Set a seed (for reproducibility)
```{r setup}
set.seed(1996)
```

Load libraries
```{r loadLibraries}

library(glmnet)
library(tidyverse)

```

## Define functions
*runGlm*: to run  multi-variant regression. This function takes a feature matrix `X` and a continuous response matrix `y`, to run lasso or ridge-regularised regression. The function will run the `cv.glmnet` function for the chosen number of `repeats`,  applying cross-fold validation using chosen number of `folds`. It outputs a list of objects.
```{r runGlm3 }
#select lasso or ridge 
runGlm <- function(X, y, method = "lasso", repeats=30, folds = 3) {
  
  #set up objects
  modelList <- list()
  lambdaList <- c()
  varExplain <- c()
  
  #set up a matrix for values of coefficients, with a row for each feature, and a column for each repeat
  coefMat <- matrix(NA, ncol(X), repeats)
  
  #make row names = genetic features
  rownames(coefMat) <- colnames(X)
  #set alpha according to selected method
  if (method == "lasso"){
    alpha = 1
  } else if (method == "ridge") {
    alpha = 0
  }
  
  #Run cv.glmnet for chosen number of repeats 
  for (i in seq(repeats)) {
    
    #if there are more than two features, fit a glm
    if (ncol(X) > 2) {
      
      res <- cv.glmnet(X,y, type.measure = "mse", family="gaussian", 
                       nfolds = folds, alpha = alpha, standardize = FALSE)
      
      #add lambda min from this repeat to the list of lambdas
      lambdaList <- c(lambdaList, res$lambda.min)
      
      #put the res object (with lambdas) into the list of models
      modelList[[i]] <- res
      
      #extract the coefficients for each feature, for lambda.min
      coefModel <- coef(res, s = "lambda.min")[-1] #remove intercept row
      
      #put these coefficients into  column of coefMatrix corresponding to the repeat
      coefMat[,i] <- coefModel
      
      #calculate variance explained
      y.pred <- predict(res, s = "lambda.min", newx = X)
      varExp <- cor(as.vector(y),as.vector(y.pred))^2
      varExplain[i] <- ifelse(is.na(varExp), 0, varExp) 
      
     
      
    } else {
      #if there are only two features, fit a linear model
      fitlm<-lm(y~., data.frame(X))
      varExp <- summary(fitlm)$r.squared
      varExplain <- c(varExplain, varExp)
      
    }
  }
  #gather all lists
  list(modelList = modelList, lambdaList = lambdaList, varExplain = varExplain, coefMat = coefMat)
}
```



## Perform regression
### Generate Feature Matrix
Set up genetic feature matrix  
```{r FeatureMatrix}
#Generate Matrix

#e.g. features from patient meta file
geneMatrix <- 
  #select mutations to include in model, make sure all features are on same scale i.e between 0-1
  dplyr::select(patMeta,c(TP53:ATM)) %>%

  #convert factors to numeric
  mutate_if(is.factor, as.character) %>%
  mutate_at(vars(-PatientID), as.numeric) %>%
  
  #convert to matrix format, with patient IDs as rownames
  data.frame() %>% 
  column_to_rownames("PatientID") %>% 
  as.matrix()

#Tidy matrix ready to use in runGlm function

#1. Remove genes with higher than 20% missing values
geneMatrix <- geneMatrix[,colSums(is.na(geneMatrix))/nrow(geneMatrix) <= 0.2]

#2. Filter for patients with complete data and no NA values (you can't input NAs)
geneMatrix.complete <- geneMatrix[complete.cases(geneMatrix),]

```

### Generate Response Matrix 
```{r responseMatrix}
#generate a reponse matrix with same dimensions as feature matrix
#e.g. generate matrix of drug response measurements for same set of patients as in genetic feature matrix
viabMatrix <- 
  #select patients in drug response data frame which are in feature matrix
  dplyr::filter(drugResponses,
                PatientID %in% row.names(geneMatrix.complete)) %>%
  
  #select drug treatment, log-transformed viability values and patient I)
  dplyr::select(treatment, 
                Log_response, 
                PatientID) %>% 
 
   #reshape data
  spread(key = PatientID, value = Log_response) %>% 
  data.frame() %>% 
  
  #make treatment the row names
  remove_rownames() %>%
  column_to_rownames("treatment")

#make order of patient samples in viability matrix the same as in feature matrix
viabMatrix <- viabMatrix[,rownames(geneMatrix.complete)]

```


### Run regression
``` {r RunRegression, warning = FALSE}

#initiate object to hold model outputs
dataResult <- list()

#e.g. fit model for each drug treatment, to identify link between drug response and genetic features
## for each drug (rowname):
for (i in rownames(viabMatrix)){  
  
    #prepare feature and response matrices
    y <- unlist(viabMatrix[i,]) # viability for all patients with this drug treatment
   
     X <- geneMatrix.complete #genetic features for all patients
    
     #fit the model
     #use lasso penalisation, to shrink coefficents and generate a sparse model with fewer coefficients
     #fit model for 30 bootstrapped repeats
    cvglmfit <- runGlm(X, y, method="lasso", repeats=30, folds=3)
    
    #collect the results for each stimulus in one object
    dataResult[[i]] <- cvglmfit
}

#save ouputs
save(dataResult, geneMatrix.complete, viabMatrix,  file = "../../data/modelOutputs.RData")
```

## Appendix
```{r appendix3}
Sys.info()
sessionInfo()
```