---
title: "Seurat_Mapping_Queries_onto_references"
author: "Holly A. R. Giles"
date: "`r doc_date()`"
output:
  BiocStyle::html_document:
      toc: yes
      toc_depth: 3
      toc_float: yes
      code_folding: "hide" 
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  fig.width = 10)
 
set.seed(1996)

```

```{r loadLibraries, include=FALSE}


library(patchwork)
library(ggplot2)
library(multtest)
library(cowplot)
library(tidyverse)
library(scDblFinder) 
library(Seurat)
library(ggplot2)
library(speckle)



```


```{r defineAesthetics, include=FALSE, warning = FALSE }

source("../../R/themes_colors.R")


```

# Introduction to single-cell reference mapping

# Reference pre-processing

## Load data
Load a pre-integrated reference

```{r}
load(file = "/data/harg2/analysis/squamousRData/controlworkspace.RData")

```

## Annotate cell types

```{r assignCellIdentity, eval = TRUE}
						 
anno.combined <- RenameIdents(libsCombined, 
                                      "0" = "", 
                                      "1" = "", 
                                      "2" = "",
                                      "3" = "",
                                      "4" = "",
                                      "5" = "", 
                                      "6" = "" , 
                                      "7" = " ", 
                                      "8" = " ",
                                      "9" = "",
                                      "10" = "",
                                      "11" = "") 


```

## Visualise UMAP

```{r cellType_annotated_UMAP, fig.path=plotDir, dev=c("png", "cairo_pdf"), eval = TRUE,  fig.height=5, fig.width=12}

namedclusterColors <-  
  list("" = "#FF0000", 
       "" = "#FFA500", 
       "" = "#00FF00",
       "" = "#008000", 
       "" = "#00FFFF", 
       "" = "#0000FF",
       " " = "#000080",
       "" = "#FF00FF", 
       "" = "#800080", 
       "" = "#FFC0CB" 
  
)


DefaultAssay(anno.combined) <- "integrated"

DimPlot(anno.combined, cols = namedclusterColors  ) 


```

# Cell type classification using an integrated reference

Seurat allows the projection of reference data (i.e. control-only) onto
a query object (i.e. treated). There are two important distinctions
between data transfer and data integration:

1.  In data transfer, Seurat does not correct or modify the query
    expression data.
2.  In data transfer, Seurat has an option (set by default) to project
    the PCA structure of a reference onto the query, instead of learning
    a joint structure with CCA. We generally suggest using this option
    when projecting data between scRNA-seq datasets.

Data transfer involves the following steps:

1\. Preprocessing the query data (in this case the 6 treated libraries)

-   For a list of Seurat objects

    -   Normalise each object, using the smae method as you did for the
        reference

    -   Find anchors for the transfer, in this case I use PCA but you
        can also use supervised PCA if you are using multimodal data
        like CITEseq

    -   Transfer cell type labels from the reference to the query.

    -   Add predictions of cell types (from the original annotation of
        the treatment data) as meta data to be able to test the accuracy
        of data transfer

2\. Merging the query and reference data

-   Project the query data onto the UMAP reference structure

-   Visualise the results for each individual library, and for all
    treated libraries together

-   Check the quality of the predictions

-   Generate a new UMAP, based on all the data, in case there are cell
    types that only exist in treated data


# Preprocessing the treated libraries 

## Normalise data

## Find transfer anchors for the transfer

## Transfer cell type labels from the reference to  query

## Add predictions of cell types 

# Merge query and reference data

##   Project the query data onto the UMAP reference structure

##   Visualise new UMAPs

### Individual

### All 

## Check the quality of the predictions

##  Generate a new UMAP for query and reference 

 

